---
title: "Blog Post 1: Exploring Past House Vote Share"
author: Jacob Moore
date: '2022-09-14'
slug: []
categories: []
tags: []
---

```{r setup, include=FALSE}
# Hiding all code output
knitr::opts_chunk$set(echo = FALSE)
# Loading libraries
library(tidyverse)
library(sf)
# Reading in data
house_vote_seat_party <- read_csv("data/house nationwide vote and seat share by party 1948-2020.csv")
house_vote_district <- read_csv("data/house party vote share by district 1948-2020.csv")

# Setup for first map
h <- house_vote_district
# filter for relevant variables
D_2020 <- h %>%
  filter(raceYear == 2020) %>%
  select(raceYear, State, district_num, district_id, RepVotes, DemVotes) %>%
# summarize party vote share by state
  group_by(State) %>%
# mutate Rep vote margin by state %>%
  mutate(R_votemargin_st = (sum(RepVotes))/
                            sum(RepVotes + DemVotes),
         D_votemargin_st = (sum(DemVotes))/
                            sum(RepVotes + DemVotes)) %>%
  rename(state = State)
# load usmap
library(usmap)
states_map <- usmap::us_map()
unique(states_map$abbr)

```

This blog post will be the first in a series of posts for Gov 1347: Election Analytics, a course at Harvard taught by Ryan Enos. Over the course of this semester, I will be building up my electoral knowledge in order to build a predictive model of the upcoming 2022 midterm elections. 

# Vote Share by State in 2020

This post will attempt to visualize past House election data to get a better sense of the past political landscape. Let's begin with the last election in 2020.

```{r visualization, warning=FALSE}
# plot
plot_usmap(data = D_2020, regions = "states", values = "D_votemargin_st") + 
scale_fill_gradient(low = "white", high = "blue", name = "Democrat two-party voteshare margin") + theme_minimal()
```
